---
title: "app-draft"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
---

---
title: "App - Interactive Dashboard"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
---

```{r setup, include=FALSE}
library(flexdashboard)
library(palmerpenguins) # what is this and do we need it for this app?
library(ggplot2)
library(dplyr)
library(leaflet) # interactive maps
library(plotly) # interactive plots
```

Page One
=======================================================================

Row 1 {data-height=600}
-----------------------------------------------------------------------

<!--When we use the orientation "rows" we specify each row using "Row".
    To adjust the size of the row use the attribute {data-height=xxx}-->

### Penguin Scatterplot

```{r}
ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
    geom_point(aes(color = species)) +
    scale_color_manual(values = c(
        "Adelie" = "purple2",
        "Chinstrap" = "orange",
        "Gentoo" = "cyan4"
    )) +
    labs(
        title = NULL,
        x = "Flipper Length (mm)",
        y = "Body Mass (g)",
        color = "Species"
    ) +
    theme_minimal()
```

Row 2
-----------------------------------------------------------------------

### Bill Depth Distribution of Chinstrap 

```{r}
penguins %>%
    filter(species == "Chinstrap") %>%
    ggplot(aes(x = bill_depth_mm)) +
    geom_histogram(fill = "orange") +
    labs(title = NULL,
         x = "Bill Depth (mm)",
         y = NULL) +
    theme_minimal()
```

### Penguin Species on Biscoe

```{r}
# subsetted penguins data frame
biscoe_sp <- penguins %>% 
    filter(island == "Biscoe") %>% 
    group_by(species) %>% 
    summarize(count = n())
    
# bar plot 
ggplot(data = biscoe_sp, aes(x = species, y = count)) +
    geom_col(aes(fill = species)) +
    scale_fill_manual(values = c(
        "Adelie" = "purple2",
        "Gentoo" = "cyan4"
    )) +
    labs(title = NULL,
         x = NULL,
         y = "Number of species",
         fill = "Species") +
    theme_minimal() 
```

Page Two
===

<!--To create a new Page or a new Row (or Column), you only need a minimum
    of three `===` (equal signs) or `---` (dashes). OR you can use a level-one     header (new Page) `#` or level-two header (new Row or Column) `##` 
    from Markdown syntax -->
    
Row {data-height=50}
---

### Total penguins

```{r}
valueBox(nrow(penguins), 
         color = "cornsilk")
```

### Penguin species

```{r}
valueBox(length(unique(penguins$species)), 
         color = "lightcyan")
```

<!-- fa = Font Awesome. You can explore free and paid font awesome icons at
    https://fontawesome.com/ -->

Row 
---

### About Palmer Station {data-width=150}

<!--To change the size of an individual component within a row (or column)
    use either {data-width} or {data-height}-->

[Palmer Station](https://pallter.marine.rutgers.edu/) is a site part of the Long-Term Ecological Research (LTER) Network. It is one of the three United States research stations located in Antarctica. 

It is also where the data from the [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/) R package was collected. The penguin data was collected and made available by [Dr. Kristen Gorman](https://www.uaf.edu/cfos/people/faculty/detail/kristen-gorman.php).

### Map

```{r}
leaflet() %>%
    addProviderTiles(providers$OpenStreetMap) %>% 
    addMarkers(lng = -64.05384063012775,
               lat = -64.77413239299318,
               popup = "Palmer Station")
```
